### 2024-12-01 paper/paper.qmd
Updated paper

---
title: "Predicting Grocery Price Trends Using Bayesian Models"
subtitle: "An analysis of price patterns across Canadian grocery vendors to identify the most beneficial vendor for consumers"
author: Tina Kim
thanks: "Code and data are available at: https://github.com/thk421/Canadian-Grocery-Prices.git"
date: today
date-format: long
abstract: "This study uses Bayesian statistical models to analyze grocery price trends across eight Canadian vendors by examining current and historical grocery price data. Our findings show that vendor-specific factors and product type significantly influence price changes. This research provides actionable insights for consumers and vendors looking to optimize their grocery purchasing and pricing strategies while adhering to the 2019 Canada's Food Guide, especially in light of the growing demand for affordable, sustainable food choices."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(brms)
library(ggplot2)
library(lubridate)
library(modelsummary)
library(arrow)

analysis_data <- read_parquet(here::here("data/02-analysis_data/analysis_data.parquet"))
```

# Introduction

Overview paragraph

The 2019 Canada's Food Guide [@cfg2019] highlights the importance of balanced nutrition, encouraging a diet abundant in vegetables, fruits, whole grains, and protein foods sourced from both plants and animals. It promotes reducing saturated fats, added sugars, and highly processed foods, while introducing greater flexibility by combining dairy products with other protein foods and recommending unsweetened, lower-fat milk or fortified plant-based alternatives. Unfortunately, it has become difficult in more recent years to adhere to the 2019 CFG and to maintain healthy eating habits due to increasing grocery prices. A basket of groceries that were once worth \$1000 five years ago costs around $1296 in 2024 [@rbc_grocery_costs]. As a result, consumers are forced to make more informed decisions when buying their groceries, especially when weighing the benefits of getting unhealthier but cheaper options and healthier but pricier options. To solve this issue, this study investigates the trends in grocery pricing across Canadian vendors, leveraging Bayesian statistical methods to model future price changes. By focusing on various predictors such as product categories, vendor information, and temporal trends, this research offers insights into the dynamics of grocery pricing.

Estimand paragraph

The key estimand of this study is the effect of vendor choice, product category, and temporal trends on grocery pricing across Canada. Specifically, we aim to estimate how much more or less consumers might pay for certain product categories when purchasing from different vendors, considering the ongoing trends in grocery pricing and the role that seasonal and economic shifts play. Through this analysis, we seek to estimate the probability of future price decreases or increases based on these factors

Results paragraph

The analysis reveals that vendor choice plays a significant role in determining grocery prices, with large national vendors generally offering lower prices compared to local or regional options. Furthermore, we observe a clear seasonal trend in grocery pricing, with prices typically rising in the winter months and dropping during certain sales periods. The model also indicates that certain product categories, particularly organic and plant-based foods, are subject to higher price volatility, while staple items such as grains and protein foods remain relatively stable.

Why it matters paragraph

Understanding how pricing trends vary across different vendors and product categories is crucial for consumers seeking to optimize their grocery spending while adhering to the dietary guidelines outlined by Canada's Food Guide. This research not only informs consumer decision-making but also provides valuable insights for policymakers and vendors aiming to address food accessibility and affordability in Canada. By modeling future price trends, we can better anticipate and prepare for fluctuations in grocery costs, helping Canadians make healthier, more cost-effective food choices.

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data describes the data used in the analysis, @sec-model details the Bayesian model and assumptions, @sec-results presents the results, and @sec-discussion offers a discussion of the findings and suggests areas for future research.

# Data {#sec-data}

## Overview

The Canadian Grocery Price Data used in this analysis paper was sourced from Project Hammer [@grocery_data]. We use the statistical programming language R [@citeR] for data processing and analysis.

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

To construct a dataset that accurately reflects the real-world phenomena we aim to study, we begin by translating observable factors into measurable entries. We focus on the pricing dynamics of grocery products from multiple vendors across Canada, which are impacted by various factors such as product type, vendor, and temporal trends.

First, the phenomenon of interest is the price of grocery products, which varies by several attributes such as vendor, product category, unit of sale, and whether the product is on promotion. We collect data from different vendors’ websites and public repositories that provide information on product names, prices, units of sale, and promotional statuses. The phenomenon of price changes over time is captured by the difference between the old and current prices, which is crucial for understanding market trends and consumer decision-making.

Once the raw data is collected, it undergoes a series of steps to ensure the quality and consistency of the dataset:

Standardization of units: Products in the dataset are sold in different units (e.g., kg, g, pack, each), which can complicate price comparisons. To address this, we standardize the units of all products to a consistent measurement, such as price per kilogram or price per liter, to allow for meaningful price comparisons across different products and vendors.

Categorization of products: We categorize products into broader groups such as “Fruits & Vegetables,” “Protein Foods,” and “Grain Products” based on keywords in the product name. This helps us understand pricing trends across different food categories and assess whether certain types of products are more susceptible to price fluctuations.

Handling missing or incorrect values: In the raw data, some records contain missing or incomplete information, such as missing prices or vendor names. These missing values are cleaned by removing rows where critical information is missing or by imputing values where applicable.

Transformations: Some variables, such as prices or units, require transformations for analysis. For instance, prices are often recorded in raw values (e.g., a price for a pack of 5 items), and we calculate the price per unit to enable more consistent analysis across items with varying quantities.

Through these steps, we systematically translate complex real-world phenomena into structured, analyzable data that can be used to answer research questions about grocery pricing dynamics.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

The primary outcome variable in our study is the price per unit of each product. This outcome variable captures the key phenomenon we are interested in: how much consumers are paying per unit of a product, after accounting for varying packaging sizes and vendor pricing strategies.

Price per Unit
The price per unit is a derived variable, calculated by dividing the current price by the unit size for each product. This allows us to standardize prices across different products that may be sold in different quantities, whether it's per kilogram, per liter, or per item. The calculation of price per unit is crucial for comparing products sold in various units of measurement and understanding the underlying cost dynamics of each product.

```{r}
#| label: tbl-price-per-unit
#| fig-cap: "Summary of price per unit across product categories and vendors"
#| echo: false
#| warning: false
#| message: false

```

Talk more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

**Temporal Information** 
   - Timestamp (nowtime): This refers to the time when the data was collected. It enables the identification of trends over time, such as the seasonal impact on stock or sales. This element is critical when applying Bayesian inference to predict future price trends. From Figure 3, we see that majority of the data was collected after October, which might bring some limitations that will be discussed further into this paper.

**Vendor Characterisics**
   - Vendor Name (vendor): There are 8 vendors in this analysis: Voila, Loblaws, NoFrills, Metro, Walmart, TandT, SaveonFoods, and Galleria. This variable is crucial since we are comparing results between these vendors. For instance, we can factor in the fact that Voila has approximately 4 times the number of products compared to Galleria as seen in the table below.

**Product Classification**
   - Food Category (food_category): Classification as “Fruits & Vegetables,” “Protein Foods,” or “Grain Products” affects weighted cost calculations. Looking at the summary table below, we see that the overall number of protein foods far outweigh grain products.

**Product Price Information**
   - Current Price (current_price): This is the most recent price of a product (in CAD) used when measuring weighted costs and quiet decreases. The table below presents the price statistics for a product range, including the minimum price (0.2), maximum price (65.99), mean price (7.03), and standard deviation (5.20). The standard deviation of 5.20 is relatively high compared to the mean price of 7.03, suggesting significant variability in prices.

   - Quiet Price Decrease (price_decrease): Tracks subtle pricing adjustments by iden- tifying instances where a product’s price decreased without being labeled as a sale. This outcome sheds light on subtle pricing strategies vendors might employ to remain com- petitive without overt promotional activities. It provides insights into whether vendors prioritize transparent discounting (sales) or discreet price reductions. As seen in Figure 4, Metro has the most number of products with decreased prices that were not on sale.

**Stock and Promotional Indicators**
   - Stock Status (stock_status): The proportion of products available (in stock) for each vendor reflects the reliability of a vendor in meeting consumer demands. Vendors with higher average stock levels may be perceived as more dependable by consumers. According to Figure 5, Loblaws has the most Fruits & Vegetables, Galleria has the most Grain Products, and Metro has the most Protein Foods.

   - On Sale (is_sale): When a product is on sale and is explicitly labeled, we can use this predictor for identifying explicit promotions that can benefit consumers. In Figure 6, we measure the average number of products explicitly labeled as “on sale” per vendor, but we notice that only Voila has this label.

   - Best Seller Status (is_best): The “Best Seller” designation is a key predictor for stock prioritization, promotional strategies, and pricing trends. As shown in Figure 7, Walmart is the only vendor that explicitly uses this label, a crucial observation to consider when developing our statistical model.

   - Organic Product Status (is_organic): Whether a product is “Organic” serves as a critical predictor for those who want higher quality products while risking affordability. We compare the availability of high-quality products across vendors and categories by looking for product names that include “fresh” or “organic”. In Figure 8, we see that Metro and Voila have the widest range of organic products. These products tend to be more pricey than normal products, which is a trade-off for people who want to invest in better quality over affordability.

# Model {#sec-model}

The model is expressed as:  
$$
\text{price\_per\_standard\_unit}_i \sim \mathcal{N}(\mu_i, \sigma^2)
$$  
where:  
- $\mu_i$ is the predicted mean price for product (i).  
- $\sigma^2$ represents the residual variance.

The mean structure is defined as:  
$$
\mu_i = \beta_0 + \beta_1 \text{nowtime}_i + \beta_2 \text{vendor}_i + \beta_3 \text{food\_category}_i + \beta_4 \text{current\_price}_i + \epsilon_i
$$

where:

- \(\beta_0\): Intercept representing the baseline price.  
- \(\text{nowtime}_i\): Continuous variable indicating the timestamp of data collection.  
- \(\text{vendor}_i\): Categorical variable with levels for the eight vendors: Voila, Loblaws, NoFrills, Metro, Walmart, TandT, SaveonFoods, and Galleria.  
- \(\text{food\_category}_i\): Categorical variable indicating "Fruits & Vegetables," "Protein Foods," or "Grain Products."  
- \(\text{current\_price}_i\): Continuous variable for the most recent price of a product.  
- \(\epsilon_i\): Random error term capturing unexplained variability.

## Priors

The Bayesian model incorporates weakly informative priors to reflect plausible ranges for parameters:  
- $\beta_0 \sim \mathcal{N}(0, 10)$: Baseline prices are expected to fall within typical ranges.  
- $\beta_1, \beta_2, \beta_3, \beta_4 \sim \mathcal{N}(0, 5)$: Moderate effect sizes are prioritized, but larger effects are not ruled out.  
- $\sigma \sim \text{Student-}t(3, 0, 10)$: Allows flexibility for variance in product pricing.

### Model Justification

The selection of predictors included or excluded in the Bayesian model was guided by their relevance, interpretability, and alignment with the research objectives. Below is a detailed breakdown:

#### Included Predictors
1. **Temporal Information**:  
   - **`nowtime`**: Captures potential seasonal trends or time-dependent price fluctuations, ensuring the model can account for temporal variability in pricing.

2. **Vendor Characteristics**:  
   - **`vendor`**: Reflects inter-vendor differences, which are critical for evaluating affordability across vendors.

3. **Product Classification**:  
   - **`food_category`**: Ensures the model incorporates the 2019 Canadian Food Guide’s proportional weightings for different food categories, linking the analysis to dietary guidelines.  
4. **Product Price Information**:  
   - **`current_price`**: Serves as the baseline price, providing essential context for predicting future prices.

These features were chosen based on insights from exploratory data analysis, ensuring that the model is interpretable and reflects the underlying structure of the dataset.

---

#### Excluded Predictors
1. **Product Price Information**:  
   - **`price_decrease`**: Highlights subtle pricing strategies, such as quiet reductions, which may not be explicitly marked as sales.

   *Reason for Exclusion*: There may be collinearity with predictors that already account for price change such as **current_price**, and may overfit due to being highly vendor-specific.

2. **Stock and Promotional Indicators**:  
   - **`stock_status`**: Represents product availability and vendor reliability.  
   - **`is_sale`**: Captures the effect of explicit promotions.  
   - **`is_best`**: Evaluates the influence of "Best Seller" labels.

   *Reason for Exclusion*:  
   These variables were excluded because they are highly vendor-specific and exhibit limited variability across vendors or over time. Including them could introduce vendor-specific biases and reduce the generalizability of the model across multiple vendors.

2. **Organic Product Status**:  
   - **`is_organic`**: Represents consumer preferences for organic products.

   *Reason for Exclusion*:  
   The impact of this variable is often reflected in other covariates, such as **`food_category`** or **`vendor`**. Including it could lead to redundancy and multicollinearity, complicating the model without providing significant additional insights.

---

This selection ensures the model is both parsimonious and robust, focusing on variables with the highest explanatory power while avoiding potential confounders or redundant predictors.

## Assumptions  
- The outcome variable, **`price_per_standard_unit`**, is normally distributed around the predicted mean.  
- Predictors have linear effects on the outcome.  
- Residual errors are independent and identically distributed.  
- Vendor and product effects are hierarchical, reflecting their nested relationships.

## Implementation and Validation  
The model was implemented in R using the `brms` package. Validation was done by visually checking posterior predictive checks to assess the model's fit.

## Limitations  
The model assumes a consistent distribution of predictors across vendors and time, which may not hold in sparse data. Additionally, the linearity assumption may oversimplify complex relationships. Future iterations could explore interaction terms or non-linear effects to enhance flexibility.

## Alternative Models Considered  
Linear regression without Bayesian inference was considered but failed to incorporate uncertainty adequately. A Bayesian logistic model was evaluated for binary outcomes like **`is_sale`**, **is_best**, and **is_organic** but could not capture the continuous nature of the primary outcome. The chosen model balances complexity and interpretability while addressing hierarchical dependencies.

# Results {#sec-results}

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| label: tbl-overall-vendor-scores
#| fig-cap: ""
#| warning: false

overall_vendor_scores_bar <-
  readRDS(file = here::here("models/overall_vendor_scores_bar.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-bayesian-model
#| fig-cap: ""
#| warning: false

bayesian_model <-
  readRDS(file = here::here("models/bayesian_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-predicted-price-trend
#| fig-cap: ""
#| warning: false

predicted_price_trend <-
  readRDS(file = here::here("models/predicted_price_trend.rds"))
```

# Discussion {#sec-discussion}

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-12-01 scripts/03-clean_data.R
Updated data cleaning

@@ -126,5 +126,4 @@ cleaned_unit_data <- cleaned_unit_data %>%
 distinct_cleaned_data <- cleaned_unit_data %>% distinct()
 
 #### Save data ####
-write_csv(distinct_cleaned_data, "data/02-analysis_data/analysis_data.csv")
 write_parquet(distinct_cleaned_data, "data/02-analysis_data/analysis_data.parquet")

### 2024-12-01 paper/paper.qmd
Updated llm usage and papers

@@
 library(lubridate)
-library(modelsummary)
 library(arrow)
+library(knitr)
 
@@
 
-The 2019 Canada's Food Guide [@cfg2019] highlights the importance of balanced nutrition, encouraging a diet abundant in vegetables, fruits, whole grains, and protein foods sourced from both plants and animals.
-It promotes reducing saturated fats, added sugars, and highly processed foods, while introducing greater flexibility by combining dairy products with other protein foods and recommending unsweetened, lower-fat milk or fortified plant-based alternatives.
-Unfortunately, it has become difficult in more recent years to adhere to the 2019 CFG and to maintain healthy eating habits due to increasing grocery prices.
-A basket of groceries that were once worth \$1000 five years ago costs around $1296 in 2024 [@rbc_grocery_costs].
-As a result, consumers are forced to make more informed decisions when buying their groceries, especially when weighing the benefits of getting unhealthier but cheaper options and healthier but pricier options.
-To solve this issue, this study investigates the trends in grocery pricing across Canadian vendors, leveraging Bayesian statistical methods to model future price changes.
-By focusing on various predictors such as product categories, vendor information, and temporal trends, this research offers insights into the dynamics of grocery pricing.
+The 2019 Canada's Food Guide (CFG) [@cfg2019] emphasizes the importance of balanced nutrition, advocating for a diet rich in vegetables, fruits, whole grains, and protein foods sourced from both plants and animals.
+It encourages reducing saturated fats, added sugars, and highly processed foods while recommending unsweetened, lower-fat milk or fortified plant-based alternatives.
+Notably, the 2019 CFG introduces greater flexibility compared to previous editions by integrating dairy products into the broader protein foods category, reflecting a more inclusive approach to nutrition.
+
+However, adhering to the 2019 CFG has become increasingly challenging due to rising grocery costs.
+A basket of groceries that cost \$1,000 five years ago now costs approximately $1,296 in 2024 [@rbc_grocery_costs].
+This escalation forces consumers to make more informed and difficult choices, often weighing the affordability of less healthy but cheaper options against the higher costs of healthier alternatives.
+
+To address this issue, our study explores grocery pricing trends across eight Canadian vendors: Metro, Save-On-Foods, Galleria, T&T Supermarket, Voilà, Walmart, No Frills, and Loblaws.
+By examining predictors such as product categories, vendor pricing, and temporal trends, this research provides critical insights into the dynamics of grocery pricing.
+Additionally, Bayesian statistical methods are employed to model future price changes, offering valuable tools for consumers and policymakers navigating the growing challenges of maintaining healthy, cost-effective diets.
 
@@
 Specifically, we aim to estimate how much more or less consumers might pay for certain product categories when purchasing from different vendors, considering the ongoing trends in grocery pricing and the role that seasonal and economic shifts play.
-Through this analysis, we seek to estimate the probability of future price decreases or increases based on these factors
+Through this analysis, we seek to estimate the probability of future price decreases or increases based on these factors.
+
+Another key estimand is the lowest total cost of constructing a week’s worth of meals using commonly available ingredients that adhere to the 2019 CFG, tailored for each vendor.
+All analysis is conducted using a standardized price measure to ensure consistency, comparability, and fairness across products, units, vendors, and contexts.
+Since prices are often reported in various units (e.g., kilograms, grams, liters, or milliliters), standardizing them to a common unit (such as grams, milliliters, or individual items) eliminates variability caused by differing measurement scales, enabling accurate and direct comparisons.
+This comparison extends beyond affordability to also consider quality, reliability, transparency, and appeal to calculate an overall score measure.
+Quality is assessed by comparing if the product is organic or not.
+Reliability is based on stock availability.
+Transparency is measured through quiet price decreases.
+Lastly, best seller or sale promotions are integrated for appeal.
+This composite metric provides a comprehensive rating for each vendor to offer tailored recommendations, helping consumers identify the vendor that best aligns with their specific needs and priorities.
 
@@
 This research not only informs consumer decision-making but also provides valuable insights for policymakers and vendors aiming to address food accessibility and affordability in Canada.
-By modeling future price trends, we can better anticipate and prepare for fluctuations in grocery costs, helping Canadians make healthier, more cost-effective food choices.
+By modeling future price trends, we can better anticipate and prepare for fluctuations in grocery costs, helping people in Canada make healthier, more cost-effective food choices.
 
@@
 
-The Canadian Grocery Price Data used in this analysis paper was sourced from Project Hammer [@grocery_data]. We use the statistical programming language R [@citeR] for data processing and analysis.
+The "Canadian Grocery Price Data" dataset was sourced from Project Hammer [@grocery_data], and a statistical analysis was performed using the R programming language [@citeR].
+The primary objective of this analysis was to clean, explore, and visualize the data to identify which grocery vendor currently and in the future offers the most cost-effective options for individuals adhering to the 2019 CFG.
+Various R libraries were utilized throughout the process.
+The tidyverse library [@tidyverse] provides a comprehensive set of tools for data manipulation, cleaning, and visualization, enabling efficient handling of datasets.
+The lubridate library [@lubridate] simplifies working with date-time objects while arrow [@arrow] is used to handle large datasets efficiently, especially for columnar formats like Parquet.
+For Bayesian modeling, rstanarm [@rstanarm] and brms [@brms] offer powerful tools for fitting and interpreting multilevel models while bayesplot [@bayesplot] aids in the graphical exploration of Bayesian model results, enhancing the interpretation of complex models.
+A summary table showcasing 10 sample observations from the cleaned dataset is presented in @tbl-grocery-data-summary.
+The table was divided into Part 1 and Part 2 for better readability, but the original dataset is formatted as one continuous table.
+
+```{r}
+#| label: tbl-grocery-data-summary
+#| tbl-cap: "Sample Canadian Grocery Price Data"
+#| echo: false
+#| warning: false
+#| message: false
+
+# First table with columns 1 to 5
+analysis_data %>%
+  slice(379030:379040) %>%
+  select(1:5) %>%  # Selecting first 5 columns
+  kable(caption = "Sample Canadian Grocery Price Data - Part 1", align = rep("c", 5))
+
+# Second table with columns 5 to the end
+analysis_data %>%
+  slice(379030:379040) %>%
+  select(6:ncol(analysis_data)) %>%  # Selecting columns 6 to the last
+  kable(caption = "Sample Canadian Grocery Price Data - Part 2", align = rep("c", ncol(analysis_data)-5))
+
+```
 
@@
 
-To construct a dataset that accurately reflects the real-world phenomena we aim to study, we begin by translating observable factors into measurable entries.
-We focus on the pricing dynamics of grocery products from multiple vendors across Canada, which are impacted by various factors such as product type, vendor, and temporal trends.
+In this analysis, the phenomenon of interest is the cost of constructing a meal plan based on the 2019 CFG by going through grocery prices of different products from various vendors.
+The original Canadian Grocery Price Data from Project Hammer used in this analysis came in two files, one containing metadata and product details, and the other containing time-series price data.
+The files were merged into one dataset and select variables were chosen to construct the analysis data for this paper.
+Detailed cleaning steps for the analysis data can be seen in @sec-appendix-B.
 
-First, the phenomenon of interest is the price of grocery products, which varies by several attributes such as vendor, product category, unit of sale, and whether the product is on promotion.
-We collect data from different vendors’ websites and public repositories that provide information on product names, prices, units of sale, and promotional statuses.
-The phenomenon of price changes over time is captured by the difference between the old and current prices, which is crucial for understanding market trends and consumer decision-making.
-
-Once the raw data is collected, it undergoes a series of steps to ensure the quality and consistency of the dataset:
-
-Standardization of units: Products in the dataset are sold in different units (e.g., kg, g, pack, each), which can complicate price comparisons.
-To address this, we standardize the units of all products to a consistent measurement, such as price per kilogram or price per liter, to allow for meaningful price comparisons across different products and vendors.
-
-Categorization of products: We categorize products into broader groups such as “Fruits & Vegetables,” “Protein Foods,” and “Grain Products” based on keywords in the product name.
-This helps us understand pricing trends across different food categories and assess whether certain types of products are more susceptible to price fluctuations.
-
-Handling missing or incorrect values: In the raw data, some records contain missing or incomplete information, such as missing prices or vendor names.
-These missing values are cleaned by removing rows where critical information is missing or by imputing values where applicable.
-
-Transformations: Some variables, such as prices or units, require transformations for analysis.
-For instance, prices are often recorded in raw values (e.g., a price for a pack of 5 items), and we calculate the price per unit to enable more consistent analysis across items with varying quantities.
-
+After the cleaning process, we are left with a dataset containing several attributes for our analysis such as vendor, product, price, and food category.
 Through these steps, we systematically translate complex real-world phenomena into structured, analyzable data that can be used to answer research questions about grocery pricing dynamics.
@@
 
-Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.
-
-The primary outcome variable in our study is the price per unit of each product.
+**Price per Unit**
+- The primary outcome variable in our study is the price per unit of each product.
 This outcome variable captures the key phenomenon we are interested in: how much consumers are paying per unit of a product, after accounting for varying packaging sizes and vendor pricing strategies.
-
-Price per Unit
-The price per unit is a derived variable, calculated by dividing the current price by the unit size for each product.
-This allows us to standardize prices across different products that may be sold in different quantities, whether it's per kilogram, per liter, or per item.
+It is calculated by dividing the current price by the unit size (in grams, milliliters, or items) for each product.
+This allows us to standardize prices across different products that may be sold in different quantities.
 The calculation of price per unit is crucial for comparing products sold in various units of measurement and understanding the underlying cost dynamics of each product.
@@
 #| label: tbl-price-per-unit
-#| fig-cap: "Summary of price per unit across product categories and vendors"
+#| tbl-cap: "Summary of price per unit across product categories and vendors"
 #| echo: false
@@
 
-Talk more about it.
+**Overall Vendor Score**
+- Another outcome variable is the overall vendor score, which takes into account multiple criteria to evaluate vendors.
+The highest weight is put on affordability which is measured by the average price per unit of groceries from each vendor.
+Some additional factors taken into account are whether products are organic (quality), are in stock (reliability), are on sale or are best sellers (appeal), and if there was a quiet price decrease (transparency).
+
+```{r}
+#| label: tbl-overall-vendor-score
+#| tbl-cap: "Summary of overall vendor scores"
+#| echo: false
+#| warning: false
+#| message: false
+
+```
+
+**Future Price Trends**
+- This outcome variable represents the predicted trajectory of grocery prices over time, estimated using a Bayesian statistical model.
+The model incorporates historical pricing data, vendor-specific trends, and product categories to generate probabilistic forecasts.
+By quantifying uncertainty, the model provides a nuanced view of potential future price fluctuations, helping consumers anticipate changes and make informed decisions.
+
+```{r}
+#| label: tbl-future-price-trends
+#| tbl-cap: "Future Price Trends"
+#| echo: false
+#| warning: false
+#| message: false
+
+```
 
 ## Predictor variables
-
-Add graphs, tables and text.
-
-Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.
 
@@
 
+```{r}
+#| label: fig-nowtime
+#| tbl-cap: "Nowtime Density Plot"
+#| echo: false
+#| warning: false
+#| message: false
+
+ggplot(analysis_data, aes(x = as.Date(nowtime))) +  # Convert to Date format for better readability
+  geom_density(fill = "lightblue", color = "black") +
+  labs(title = "Density of Data Collection Dates", x = "Date", y = "Density") +
+  theme_minimal()
+
+```
+
 **Vendor Characterisics**
@@
 
+```{r}
+#| label: tbl-vendors
+#| tbl-cap: "Summary table for vendors"
+#| echo: false
+#| warning: false
+#| message: false
+
+# Summary table for Vendor Names
+vendor_summary <- analysis_data %>%
+  count(vendor) %>%
+  arrange(desc(n))  # Sorting by count, descending
+
+# Display the table
+knitr::kable(vendor_summary, caption = "Summary of Vendor Names", col.names = c("Vendor", "Product Count"))
+
+```
+
 **Product Classification**
@@
 Looking at the summary table below, we see that the overall number of protein foods far outweigh grain products.
+
+```{r}
+#| label: tbl-food-category
+#| tbl-cap: "Summary table for food category"
+#| echo: false
+#| warning: false
+#| message: false
+
+# Summary table for Food Category
+category_summary <- analysis_data %>%
+  count(food_category) %>%
+  arrange(desc(n))  # Sorting by count, descending
+
+# Display the table
+knitr::kable(category_summary, caption = "Summary of Food Categories", col.names = c("Food Category", "Product Count"))
+
+```
 
@@
 
-   - Quiet Price Decrease (price_decrease): Tracks subtle pricing adjustments by iden- tifying instances where a product’s price decreased without being labeled as a sale.
-This outcome sheds light on subtle pricing strategies vendors might employ to remain com- petitive without overt promotional activities.
+```{r}
+#| label: tbl-current-price
+#| tbl-cap: "Summary table for current price"
+#| echo: false
+#| warning: false
+#| message: false
+
+# Summary table for Current Price
+price_summary <- analysis_data %>%
+  summarise(
+    Min_Price = min(current_price, na.rm = TRUE),
+    Max_Price = max(current_price, na.rm = TRUE),
+    Mean_Price = mean(current_price, na.rm = TRUE),
+    SD_Price = sd(current_price, na.rm = TRUE)
+  )
+
+# Display the table
+knitr::kable(price_summary, caption = "Summary of Current Prices", col.names = c("Min Price", "Max Price", "Mean Price", "Standard Deviation"))
+
+```
+
+- Quiet Price Decrease (price_decrease): Tracks subtle pricing adjustments by identifying instances where a product’s price decreased without being labeled as a sale.
+This outcome sheds light on subtle pricing strategies vendors might employ to remain competitive without overt promotional activities.
 It provides insights into whether vendors prioritize transparent discounting (sales) or discreet price reductions.
-As seen in Figure 4, Metro has the most number of products with decreased prices that were not on sale.
+As seen in @fig-price-decrease, Metro has the most number of products with decreased prices that were not on sale.
+
+```{r}
+#| label: fig-price-decrease
+#| fig-cap: "Product quiet price decrease"
+#| echo: false
+#| warning: false
+#| message: false
+
+quiet_price_dec_bar <-
+  readRDS(file = here::here("models/quiet_price_dec_bar.rds"))
+
+```
 
@@
 Vendors with higher average stock levels may be perceived as more dependable by consumers.
-According to Figure 5, Loblaws has the most Fruits & Vegetables, Galleria has the most Grain Products, and Metro has the most Protein Foods.
+According to @fig-stock-status, Loblaws has the most Fruits & Vegetables, Galleria has the most Grain Products, and Metro has the most Protein Foods.
+
+```{r}
+#| label: fig-stock-status
+#| fig-cap: "Product stock status"
+#| echo: false
+#| warning: false
+#| message: false
+
+avg_stock_facet <-
+  readRDS(file = here::here("models/avg_stock_facet.rds"))
+
+```
 
 - On Sale (is_sale): When a product is on sale and is explicitly labeled, we can use this predictor for identifying explicit promotions that can benefit consumers.
-In Figure 6, we measure the average number of products explicitly labeled as “on sale” per vendor, but we notice that only Voila has this label.
+In @fig-on-sale, we measure the average number of products explicitly labeled as “on sale” per vendor, but we notice that only Voila has this label.
+
+```{r}
+#| label: fig-on-sale
+#| fig-cap: "Number of products on sale"
+#| echo: false
+#| warning: false
+#| message: false
+
+is_sale_bar <-
+  readRDS(file = here::here("models/is_sale_bar.rds"))
+
+```
 
 - Best Seller Status (is_best): The “Best Seller” designation is a key predictor for stock prioritization, promotional strategies, and pricing trends.
-As shown in Figure 7, Walmart is the only vendor that explicitly uses this label, a crucial observation to consider when developing our statistical model.
+As shown in @fig-best-seller, Walmart is the only vendor that explicitly uses this label, a crucial observation to consider when developing our statistical model.
+
+```{r}
+#| label: fig-best-seller
+#| fig-cap: "Number of best-seller products"
+#| echo: false
+#| warning: false
+#| message: false
+
+is_best_bar <-
+  readRDS(file = here::here("models/is_best_bar.rds"))
+
+```
 
@@
 These products tend to be more pricey than normal products, which is a trade-off for people who want to invest in better quality over affordability.
+
+```{r}
+#| label: fig-is-organic
+#| fig-cap: "Number of organic products"
+#| echo: false
+#| warning: false
+#| message: false
+
+organic_product_bar <-
+  readRDS(file = here::here("models/organic_product_bar.rds"))
+
+```
 
@@
 
-# Appendix {-}
+# Appendix: Sampling Methdology Overview - Weaknesses and Potential Improvements {#sec-appendix-A}
 
-# Additional data details
+# Appendix: Additional Data Details {#sec-appendix-B}
 
-# Model details {#sec-model-details}
+## Data Cleaning Steps
+
+To prepare the dataset for analysis, we conducted comprehensive data cleaning.
+First, product and price datasets were merged based on product identifiers to create a unified dataset.
+Rows with missing critical information, such as unit or price, were removed.
+To standardize product categorization, keywords were defined for three primary food categories: fruits and vegetables, protein foods, and grain products.
+Products matching these keywords were assigned to their respective categories, while others were excluded.
+
+Unit standardization was implemented to ensure consistency in price comparisons.
+Units were cleaned, converted to lowercase, and mapped to standardized units (e.g., grams, milliliters).
+A conversion table was applied to normalize all prices to a standard per-unit measure.
+Additional flags were created to capture stock status, sale indicators, best-seller labels, and organic products based on product descriptions.
+
+Unnecessary columns, such as URLs, brands, and redundant identifiers, were dropped to streamline the dataset.
+Prices were then adjusted to calculate standardized price-per-unit values, enabling direct comparisons across products and vendors.
+Rows with unrecognized units or missing standardized price information were filtered out.
+Finally, duplicate entries were removed to ensure data integrity, and the cleaned dataset was sorted by vendor for further analysis.
+
+# Appendix: Additional Model Details {#sec-appendix-C}
 
@@
 
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
 ```
@@
 
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
 ```

### 2024-12-02 scripts/06-model_data.R
Updated model script

@@ -51,9 +51,9 @@ overall_best_vendor_data <- weekly_meal_plan_cost_per_vendor %>%
     promotional_effort = normalize(avg_products_on_sale),  # Higher sale frequency is better
     
     # overall score calculation
-    overall_score = 0.4 * affordability +
+    overall_score = 0.5 * affordability +
       0.2 * reliability +
-      0.2 * sale_transparency +
+      0.1 * sale_transparency +
       0.1 * product_appeal +
       0.1 * promotional_effort
   ) %>%

### 2024-12-02 paper/paper.qmd
Updated paper

@@
 library(knitr)
+library(modelsummary)
+library(rstan)
 
@@
 
-# First table with columns 1 to 5
+# First table with columns 1 to 4
 analysis_data %>%
-  slice(379030:379040) %>%
+  slice(370430:370440) %>%
   select(1:5) %>%  # Selecting first 5 columns
@@
 analysis_data %>%
-  slice(379030:379040) %>%
+  slice(370430:370440) %>%
   select(6:ncol(analysis_data)) %>%  # Selecting columns 6 to the last
@@
 **Price per Unit**
+
 - The primary outcome variable in our study is the price per unit of each product.
@@
 It is calculated by dividing the current price by the unit size (in grams, milliliters, or items) for each product.
-This allows us to standardize prices across different products that may be sold in different quantities.
+This can be seen in @tbl-price-per-unit, where the current price of the 'Portion Of Frozen Atlantic Salmon' product is standardized from 2.99 to 0.00299 to match the standardized unit, while the other products in the table remain the same because they are already standardized.
 The calculation of price per unit is crucial for comparing products sold in various units of measurement and understanding the underlying cost dynamics of each product.
@@
 
+# Set random seed for reproducibility
+set.seed(100)
+
+# Create a summary table with selected columns and sample 5 random observations
+summary_table <- analysis_data %>%
+  select(product_name, current_price, price_per_standard_unit) %>%
+  sample_n(5)  # Sample 5 random rows
+
+# Display the summary table
+kable(summary_table, caption = "Summary Table of Product Prices")
+
 ```
@@
 **Overall Vendor Score**
-- Another outcome variable is the overall vendor score, which takes into account multiple criteria to evaluate vendors.
+
+- Another outcome variable is the overall vendor score as seen in @fig-overall-vendor-score, which takes into account multiple criteria to evaluate vendors.
 The highest weight is put on affordability which is measured by the average price per unit of groceries from each vendor.
@@
 ```{r}
-#| label: tbl-overall-vendor-score
-#| tbl-cap: "Summary of overall vendor scores"
+#| label: fig-overall-vendor-score
+#| fig-cap: "Summary of overall vendor scores"
 #| echo: false
@@
 
+overall_vendor_scores_bar <-
+  readRDS(file = here::here("models/overall_vendor_scores_bar.rds"))
+
+overall_vendor_scores_bar
 ```
@@
 **Future Price Trends**
-- This outcome variable represents the predicted trajectory of grocery prices over time, estimated using a Bayesian statistical model.
+
+- This outcome variable as seen in @fig-future-price-trends represents the predicted trajectory of grocery prices over time, estimated using a Bayesian statistical model.
 The model incorporates historical pricing data, vendor-specific trends, and product categories to generate probabilistic forecasts.
@@
 ```{r}
-#| label: tbl-future-price-trends
-#| tbl-cap: "Future Price Trends"
+#| label: fig-future-price-trends
+#| fig-cap: "Predicted Future Price Trends"
 #| echo: false
@@
 
+predicted_price_trend <-
+  readRDS(file = here::here("models/predicted_price_trend.rds"))
+
+predicted_price_trend
 ```
@@
 **Temporal Information**
+
 - Timestamp (nowtime): This refers to the time when the data was collected.
@@
 This element is critical when applying Bayesian inference to predict future price trends.
-From Figure 3, we see that majority of the data was collected after October, which might bring some limitations that will be discussed further into this paper.
+From @fig-nowtime, we see that majority of the data was collected after October, which might bring some limitations that will be discussed further into this paper.
 
@@
 #| label: fig-nowtime
-#| tbl-cap: "Nowtime Density Plot"
+#| fig-cap: "Nowtime Density Plot"
 #| echo: false
@@
 **Vendor Characterisics**
+
 - Vendor Name (vendor): There are 8 vendors in this analysis: Voila, Loblaws, NoFrills, Metro, Walmart, TandT, SaveonFoods, and Galleria.
@@
 **Product Classification**
+
 - Food Category (food_category): Classification as “Fruits & Vegetables,” “Protein Foods,” or “Grain Products” affects weighted cost calculations.
@@
 **Product Price Information**
+
 - Current Price (current_price): This is the most recent price of a product (in CAD) used when measuring weighted costs and quiet decreases.
@@
 #| label: fig-price-decrease
-#| fig-cap: "Product quiet price decrease"
+#| fig-cap: "Count of quiet price decreases by vendor"
 #| echo: false
@@
 
+quiet_price_dec_bar
 ```
@@
 **Stock and Promotional Indicators**
+
 - Stock Status (stock_status): The proportion of products available (in stock) for each vendor reflects the reliability of a vendor in meeting consumer demands.
@@
 #| label: fig-stock-status
-#| fig-cap: "Product stock status"
+#| fig-cap: "Count of products in stock by vendor and food category"
 #| echo: false
@@
 
+avg_stock_facet
 ```
@@
 #| label: fig-on-sale
-#| fig-cap: "Number of products on sale"
+#| fig-cap: "Sale occurrence by vendor"
 #| echo: false
@@
 
+is_sale_bar
 ```
@@
 #| label: fig-best-seller
-#| fig-cap: "Number of best-seller products"
+#| fig-cap: "Proportion of best sellers by vendor"
 #| echo: false
@@
 
+is_best_bar
 ```
@@
 We compare the availability of high-quality products across vendors and categories by looking for product names that include “fresh” or “organic”.
-In Figure 8, we see that Metro and Voila have the widest range of organic products.
+In @fig-is-organic, we see that Metro and Voila have the widest range of organic products.
 These products tend to be more pricey than normal products, which is a trade-off for people who want to invest in better quality over affordability.
@@
 #| label: fig-is-organic
-#| fig-cap: "Number of organic products"
+#| fig-cap: "Count of organic products by vendor and food category"
 #| echo: false
@@
 
+organic_product_bar
 ```
@@
 
-- \(\beta_0\): Intercept representing the baseline price.  
-- \(\text{nowtime}_i\): Continuous variable indicating the timestamp of data collection.  
-- \(\text{vendor}_i\): Categorical variable with levels for the eight vendors: Voila, Loblaws, NoFrills, Metro, Walmart, TandT, SaveonFoods, and Galleria.  
-- \(\text{food\_category}_i\): Categorical variable indicating "Fruits & Vegetables," "Protein Foods," or "Grain Products."  
-- \(\text{current\_price}_i\): Continuous variable for the most recent price of a product.  
-- \(\epsilon_i\): Random error term capturing unexplained variability.
+- $\beta_0$: Intercept representing the baseline price.  
+- $\text{nowtime}_i$: Continuous variable indicating the timestamp of data collection.  
+- $\text{vendor}_i$: Categorical variable with levels for the eight vendors: Voila, Loblaws, NoFrills, Metro, Walmart, TandT, SaveonFoods, and Galleria.  
+- $\text{food\_category}_i$: Categorical variable indicating "Fruits & Vegetables," "Protein Foods," or "Grain Products."  
+- $\text{current\_price}_i$: Continuous variable for the most recent price of a product.  
+- $\epsilon_i$: Random error term capturing unexplained variability.
 
@@
 
-The selection of predictors included or excluded in the Bayesian model was guided by their relevance, interpretability, and alignment with the research objectives. Below is a detailed breakdown:
+The selection of predictors included or excluded in the Bayesian model was guided by their relevance, interpretability, and alignment with the research objectives.
+A detailed breakdown can be seen in @sec-appendix-C.
 
-#### Included Predictors
+## Assumptions  
+- The outcome variable, **`price_per_standard_unit`**, is normally distributed around the predicted mean.  
+- Predictors have linear effects on the outcome.  
+- Residual errors are independent and identically distributed.  
+- Vendor and product effects are hierarchical, reflecting their nested relationships.
+
+## Implementation and Validation  
+The model was implemented in R using the `brms` package. Validation was done by visually checking posterior predictive checks to assess the model's fit.
+
+## Limitations  
+The model assumes a consistent distribution of predictors across vendors and time, which may not hold in sparse data.
+Additionally, the linearity assumption may oversimplify complex relationships.
+Future iterations could explore interaction terms or non-linear effects to enhance flexibility.
+
+## Alternative Models Considered  
+Linear regression without Bayesian inference was considered but failed to incorporate uncertainty adequately.
+A Bayesian logistic model was evaluated for binary outcomes like **`is_sale`**, **is_best**, and **is_organic** but could not capture the continuous nature of the primary outcome.
+The chosen model balances complexity and interpretability while addressing hierarchical dependencies.
+
+# Results {#sec-results}
+
+The average weekly meal plan cost by vendor and food category in @fig-avg-price-stacked shows in order which vendors overall are the most cost-effective to the least cost-effective.
+By looking at the different colors that represent the food catgory, it can also be compared which food group is more cost-effective in which store.
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: fig-avg-price-stacked
+#| fig-cap: "Average weekly meal plan cost by vendor and food category"
+#| warning: false
+
+avg_price_stacked_bar <-
+  readRDS(file = here::here("models/avg_price_stacked_bar.rds"))
+
+avg_price_stacked_bar
+```
+
+The overall best vendor can be seen in @fig-overall-vendor-score, and the predicted future price trends per vendor is shown in @fig-future-price-trends.
+
+Our Bayesian model results are summarized in @tbl-model-summary.
+It provides the parameter estimates, predictive metrics, and significance levels to support the model.
+Further discussion for these results can be found in @sec-second-point.
+
+```{r}
+#| echo: false
+#| eval: true
+#| label: tbl-model-summary
+#| tbl-cap: "Bayesian model for predicted future price trends summary"
+#| warning: false
+
+bayesian_model <-
+  readRDS(file = here::here("models/bayesian_model.rds"))
+
+# Generate a summary table for the Bayesian model
+model_table <- modelsummary(
+  bayesian_model, 
+  stars = TRUE,
+  coef_omit = "(Intercept)"
+)
+
+# Print the table
+model_table
+```
+
+# Discussion {#sec-discussion}
+
+If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+
+## Paper Summary {#sec-first-point}
+what was done in this paper?
+
+## Bayesian Model Explanation {#sec-second-point}
+
+### Key Terms
+
+- b_current_price (0.955): The regression coefficient for current_price.
+This indicates that for each unit increase in current_price, the future price is expected to increase by 0.955 units, holding other variables constant.
+
+- b_food_categoryGrainProducts (0.065): Grain products are associated with a slight positive increase in the predicted future price relative to the baseline category (likely "Fruits & Vegetables").
+
+- b_food_categoryProteinFoods (0.178): Protein foods are associated with a larger increase in the predicted future price compared to the baseline category.
+
+- b_vendor (e.g., -0.013 for Loblaws):* These coefficients represent the effect of each vendor relative to the reference vendor.
+For instance, Loblaws has a slight negative impact on future prices compared to the baseline vendor.
+
+- b_month (0.008): The slight positive effect of the month variable suggests minor upward trends in prices over time.
+
+- sigma (1.596): This represents the residual standard deviation, indicating the variability of the outcome variable (future price) that is not explained by the predictors.
+
+- R² (0.903): This high value indicates that 90.3% of the variance in the future price is explained by the predictors in the model.
+
+- R² Adj. (0.903): Adjusted R² accounts for the number of predictors and indicates a similarly strong fit.
+
+- R² Marg. (0.903): Marginal R² reflects the variance explained by fixed effects only.
+
+- ICC (0.0): The Intraclass Correlation Coefficient suggests no substantial variance at the grouping level (e.g., product name in this case).
+
+### Predictive Metrics
+
+- ELPD (-9361.9): The Expected Log Predictive Density is a measure of predictive accuracy, where higher (less negative) values indicate better predictions.
+
+- ELPD s.e. (338.1): The standard error for ELPD, showing uncertainty around this estimate.
+
+- LOOIC (18,723.8): The Leave-One-Out Information Criterion measures model fit; smaller values are better. It is often used for model comparison.
+
+- LOOIC s.e. (676.2): The standard error of the LOOIC estimate.
+
+- WAIC (18,729.2): The Watanabe-Akaike Information Criterion is another metric for model fit, similar to LOOIC. Again, smaller values are better.
+
+- RMSE (1.59): The Root Mean Square Error quantifies the average error of the model's predictions, with lower values indicating better fit.
+
+### Bayesian Model Results Summary
+
+The high R² and R² Marg.
+values suggest the model explains most of the variability in future prices.
+Amongst the predictors, **current_price** is the strongest predictor, with nearly a 1:1 relationship to future prices.
+Other variables like **food categories** and **vendors** have smaller but notable effects.
+Metrics like LOOIC and RMSE show that the model performs well, but further investigation into predictive accuracy and residual patterns could refine interpretations.
+
+### Predicted Future Price Trends
+
+According to the predicted future price trends in @fig-future-price-trends, Galleria exhibits relatively stable predicted price trends, with slight variations over time.
+The predicted prices stay consistent, without significant spikes, suggesting steady pricing strategies.
+Loblaws shows stable prices with slight fluctuations but no dramatic increases or decreases.
+This implies consistent pricing practices with minor seasonal or product-related variations.
+Metro displays a similar pattern of stability to Loblaws but with marginally larger variations.
+The periodicity of the vertical bars may indicate uncertainty in the predictions during certain months.
+NoFrills has a noticeable spike mid-period, followed by a return to stability.
+This indicates a potential pricing anomaly or predicted fluctuation in a specific month before returning to normal pricing levels.
+SaveOnFoods presents consistent prices with some minor upward shifts, suggesting a gradual increase or variability in certain product categories.
+However, sporadic trends can be seen where there is an abrupt jump between months, which could reflect noise in the data.
+T&T has a slight peak mid-period, followed by stabilization.
+This may reflect a seasonal effect or a one-time price fluctuation for specific products.
+Voila shows a small spike mid-period but generally maintains consistent prices.
+The predicted pricing pattern mirrors a temporary fluctuation, likely tied to vendor or product-specific dynamics.
+Walmart demonstrates the most volatility in its price trends, with noticeable spikes.
+This may highlight significant uncertainty in Walmart's pricing patterns or irregularities caused by promotional events or external market forces.
+The vertical lines represent the uncertainty intervals (credible intervals) of the Bayesian predictions.
+Wider intervals imply greater uncertainty about the predicted prices for that month.
+These intervals can stem from limited data availability for specific vendors or products, variability in historical pricing trends, or external factors influencing the model’s predictive accuracy.
+
+## Overall Best Vendor {#sec-third-point}
+
+Choosing the best vendor based on the predicted future price trends depends on the specific priorities of consumers, such as affordability, price stability, and certainty.
+For affordability, NoFrills appears to have the lowest predicted prices overall, despite a noticeable spike in one period.
+Its consistent low prices outside this anomaly make it the most affordable option for price-sensitive consumers.
+Walmart also shows relatively low predicted prices but with more variability and uncertainty than NoFrills.
+For price stability, Loblaws and Metro exhibit the most stable price trends, with minimal fluctuations or uncertainty over time.
+For consumers prioritizing predictability in costs, these vendors may be the best choice.
+Galleria, Loblaws, and Metro have narrower credible intervals (vertical bars), indicating higher confidence in the model’s predictions for their prices.
+This suggests these vendors may have more predictable pricing strategies, making them reliable options.
+On the other hand, Walmart and NoFrills have wider credible intervals, reflecting more uncertainty in their pricing.
+This could indicate a higher risk of price volatility.
+Thus, based on the bayesian model, we conclude that NoFrills emerges as the best option when considering affordability, as its prices are consistently low with only one significant spike.
+Loblaws or Metro could be the best option for consumers who value price stability and predictability over absolute cost.
+
+Taking into account the Overall Vendor Score from @fig-overall-vendor-score, we can now make a more comprehensive evaluation.
+We observe that Voila achieves the highest rating, while Galleria ranks the lowest.
+These scores were primarily driven by affordability, which accounted for 50% of the total rating.
+Other factors contributing to the score included reliability (20%), measured by the vendor's consistency in keeping products in stock; sale transparency (10%), reflecting whether the vendor avoids quiet price increases; product appeal (10%), with a focus on the availability of organic products; and promotional efforts (10%), which evaluated how clearly products were marketed as being on sale.
+Despite its higher prices compared to others, Voila has the highest overall vendor score.
+This high score suggests that it performs well across multiple factors like reliability, transparency, product appeal, and promotional efforts.
+These qualities could make it the best choice for consumers who prioritize a combination of affordability and additional factors like product quality, transparency, and sales efforts.
+Voila seems to provide strong value beyond just price, appealing to consumers seeking a well-rounded vendor.
+Although NoFrills has the second-best overall vendor score, it is the most affordable in terms of predicted prices.
+This makes it an excellent choice for consumers primarily focused on low cost.
+However, it ranks slightly lower in the overall score because factors like transparency and promotional efforts are less emphasized compared to Voila.
+
+Given the importance of both affordability and vendor quality (including reliability, appeal, and transparency), Voila should be considered the best vendor.
+Despite higher prices compared to NoFrills, Voila offers a balanced combination of affordable pricing and superior service, making it the most well-rounded option overall.
+It also has narrower credible intervals in the predicted price trends compared to NoFrills, making it a more stable choice.
+
+## Weaknesses {#sec-fourth-point}
+
+While this study provides valuable insights into the dynamics of grocery pricing across different vendors, there are several limitations worth considering.
+First, the model's reliance on historical pricing data and vendor-specific trends may not fully capture the complexity of real-world pricing fluctuations, which are influenced by a variety of external factors such as market shifts, supply chain disruptions, or changes in consumer behavior.
+Furthermore, the data collected might not show accurate prediction of price tends throughout the year because most of the data collected as seen in @fig-nowtime is from after July.
+As a result, while the predictions offer valuable insights, they may not always align with actual price trends in the future.
+
+Second, the analysis primarily focused on a few key variables (such as current price, vendor, and food category), and may have overlooked other potentially important factors.
+For example, changes in consumer preferences, seasonality, and macroeconomic conditions could have significant impacts on prices and may not have been fully integrated into the model.
+This represents a gap that could be explored in future studies.
+
+Additionally, the dataset was limited to a specific time frame (2019-2020) and geographic region (Canada), meaning that the findings may not be generalizable to other countries or time periods.
+To improve the model’s applicability, future research could expand the dataset to include a broader range of time periods, geographic regions, and product categories.
+
+## Next Steps {#sec-fifth-point}
+
+To address these weaknesses, there are several avenues for future research.
+First, incorporating more granular data, such as real-time pricing or additional product attributes (e.g., packaging size, brand loyalty, or consumer ratings), could provide a more comprehensive understanding of price dynamics and vendor performance.
+Expanding the model to include additional variables could lead to more accurate predictions and help better capture the real-world complexities of grocery pricing.
+
+Second, future studies could explore the impact of external factors such as supply chain disruptions, inflation, or environmental sustainability initiatives on grocery prices.
+This would involve integrating external data sources to complement the existing dataset and help contextualize the price trends observed in this study.
+
+Lastly, conducting similar analyses across different countries or regions would allow for a comparative assessment of pricing strategies and vendor performance, providing valuable insights for international policy makers, researchers, and consumers.
+This could also facilitate the development of more robust, generalized pricing models that can be applied across various markets.
+
+\newpage
+
+\appendix
+
+# Appendix: Sampling Methdology Overview - Weaknesses and Potential Improvements {#sec-appendix-A}
+
+The Canadian Grocery Price Data sourced from Project Hammer provides valuable insights into grocery pricing trends across vendors, product categories, and regions.
+The data was obtained through a screen scrape of the website's user interface, which is why it lacks certain details that would typically be available through the internal APIs powering the grocers' websites.
+However, the sampling methodology underlying this dataset presents several weaknesses that could impact the robustness and generalizability of findings.
+
+## Weaknesses
+
+1.
+The dataset was limited to a specific time frame (2019-2020) and geographic region (Canada), meaning that the findings may not be generalizable to other countries or time periods.
+Specifically, data is collected starting from February 28, 2024, but these were for smaller baskets of products.
+Majority of the usable data with a variety of products starts from July, potentially missing seasonal variations or short-term fluctuations in prices caused by external factors such as supply chain disruptions, weather events, or promotions.
+
+2.
+The dataset relies on product identifiers and descriptions to standardize items across vendors.
+Inconsistent labeling or categorization may lead to mismatches or incomplete comparisons, particularly for highly variable products like produce.
+
+3.
+There may also be missing data for certain days for certain vendors when specific extracts failed.
+Instances of missing unit information, product details, or pricing history reduce the completeness of the dataset and necessitate assumptions or imputation that may introduce errors.
+
+## Improvements
+
+While the current sampling methodology provides valuable insights, several improvements can enhance the dataset's robustness, accuracy, and overall usefulness for analysis:
+
+1.
+To capture a more comprehensive picture of pricing trends, data collection should span a longer time frame, ideally including multiple seasons and holidays.
+This would help account for seasonal variations, promotional periods, and other temporal factors influencing grocery prices.
+A broader time range would allow for a better understanding of price fluctuations throughout the year, enabling more accurate predictions and trend analysis.
+
+2.
+Implement a more refined product matching system, potentially by using standardized product IDs or barcodes across vendors.
+Additionally, categorization should be standardized across all products to ensure accurate comparisons, particularly for perishable goods like produce.
+This would reduce mismatches and make comparisons between vendors more reliable, ensuring that price trends reflect genuine product similarities rather than differences in labeling or categorization.
+
+3.
+Enhance the data collection process by improving the scraping method to reduce data loss.
+Using a more robust scraping framework that retries failed extractions or integrates better error handling can reduce instances of missing data.
+This would improve the completeness of the dataset, ensuring a more reliable foundation for analysis and reducing the need for imputation that might introduce errors.
+Where possible, supplement the screen-scraped data with data sourced from the internal APIs of the grocers' websites.
+Internal APIs tend to offer more accurate, comprehensive, and real-time data, including product details, stock levels, and historical price trends.
+This would increase the accuracy of the dataset and provide additional insights into product availability, pricing trends, and promotions, leading to more precise analyses.
+
+# Appendix: Additional Data Details {#sec-appendix-B}
+
+## Data Cleaning Steps
+
+To prepare the dataset for analysis, we conducted comprehensive data cleaning.
+First, product and price datasets were merged based on product identifiers to create a unified dataset.
+Rows with missing critical information, such as unit or price, were removed.
+To standardize product categorization, keywords were defined for three primary food categories: fruits and vegetables, protein foods, and grain products.
+Products matching these keywords were assigned to their respective categories, while others were excluded.
+
+Unit standardization was implemented to ensure consistency in price comparisons.
+Units were cleaned, converted to lowercase, and mapped to standardized units (e.g., grams, milliliters).
+A conversion table was applied to normalize all prices to a standard per-unit measure.
+Additional flags were created to capture stock status, sale indicators, best-seller labels, and organic products based on product descriptions.
+
+Unnecessary columns, such as URLs, brands, and redundant identifiers, were dropped to streamline the dataset.
+Prices were then adjusted to calculate standardized price-per-unit values, enabling direct comparisons across products and vendors.
+Rows with unrecognized units or missing standardized price information were filtered out.
+Finally, duplicate entries were removed to ensure data integrity, and the cleaned dataset was sorted by vendor for further analysis.
+
+# Appendix: Additional Model Details {#sec-appendix-C}
+
+## Included Predictors in the Bayesian Model
 1. **Temporal Information**:  
@@
 
-#### Excluded Predictors
+## Excluded Predictors in the Bayesian Model
 1. **Product Price Information**:  
@@
 
-## Assumptions  
-- The outcome variable, **`price_per_standard_unit`**, is normally distributed around the predicted mean.  
-- Predictors have linear effects on the outcome.  
-- Residual errors are independent and identically distributed.  
-- Vendor and product effects are hierarchical, reflecting their nested relationships.
-
-## Implementation and Validation  
-The model was implemented in R using the `brms` package. Validation was done by visually checking posterior predictive checks to assess the model's fit.
-
-## Limitations  
-The model assumes a consistent distribution of predictors across vendors and time, which may not hold in sparse data.
-Additionally, the linearity assumption may oversimplify complex relationships.
-Future iterations could explore interaction terms or non-linear effects to enhance flexibility.
-
-## Alternative Models Considered  
-Linear regression without Bayesian inference was considered but failed to incorporate uncertainty adequately.
-A Bayesian logistic model was evaluated for binary outcomes like **`is_sale`**, **is_best**, and **is_organic** but could not capture the continuous nature of the primary outcome.
-The chosen model balances complexity and interpretability while addressing hierarchical dependencies.
-
-# Results {#sec-results}
-
-Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-overall-vendor-scores
-#| fig-cap: ""
-#| warning: false
-
-overall_vendor_scores_bar <-
-  readRDS(file = here::here("models/overall_vendor_scores_bar.rds"))
-```
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-bayesian-model
-#| fig-cap: ""
-#| warning: false
-
-bayesian_model <-
-  readRDS(file = here::here("models/bayesian_model.rds"))
-```
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-predicted-price-trend
-#| fig-cap: ""
-#| warning: false
-
-predicted_price_trend <-
-  readRDS(file = here::here("models/predicted_price_trend.rds"))
-```
-
-# Discussion {#sec-discussion}
-
-## First discussion point {#sec-first-point}
-
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
-
-## Second discussion point
-
-Please don't use these as sub-heading labels - change them to be what your point actually is.
-
-## Third discussion point
-
-## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
-
-\newpage
-
-\appendix
-
-# Appendix: Sampling Methdology Overview - Weaknesses and Potential Improvements {#sec-appendix-A}
-
-# Appendix: Additional Data Details {#sec-appendix-B}
-
-## Data Cleaning Steps
-
-To prepare the dataset for analysis, we conducted comprehensive data cleaning.
-First, product and price datasets were merged based on product identifiers to create a unified dataset.
-Rows with missing critical information, such as unit or price, were removed.
-To standardize product categorization, keywords were defined for three primary food categories: fruits and vegetables, protein foods, and grain products.
-Products matching these keywords were assigned to their respective categories, while others were excluded.
-
-Unit standardization was implemented to ensure consistency in price comparisons.
-Units were cleaned, converted to lowercase, and mapped to standardized units (e.g., grams, milliliters).
-A conversion table was applied to normalize all prices to a standard per-unit measure.
-Additional flags were created to capture stock status, sale indicators, best-seller labels, and organic products based on product descriptions.
-
-Unnecessary columns, such as URLs, brands, and redundant identifiers, were dropped to streamline the dataset.
-Prices were then adjusted to calculate standardized price-per-unit values, enabling direct comparisons across products and vendors.
-Rows with unrecognized units or missing standardized price information were filtered out.
-Finally, duplicate entries were removed to ensure data integrity, and the cleaned dataset was sorted by vendor for further analysis.
-
-# Appendix: Additional Model Details {#sec-appendix-C}
-
 ## Posterior predictive check
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+In @fig-pp-check, we implement a posterior predictive check for the bayesian model of our analysis.
+This shows an overlay of the observed data density (y) and the posterior predictive densities (y_rep) drawn from the model.
+The dark line represents the observed data density (y), while the lighter lines represent the posterior predictive densities (y_rep) simulated from the model.
+The predictive densities align well with the observed density in the bulk of the distribution (around 0 to 10), indicating that the model is capturing the central tendency of the data reasonably well.
+There appears to be a slight discrepancy in the right tail of the distribution (beyond ~10).
+The observed density drops more steeply than most of the predictive densities, suggesting that the model may be slightly overestimating the probability of higher prices.
+Overall, the model fits the main part of the data distribution adequately and the slight deviation in the tail may be acceptable, but if predicting high prices accurately is critical, the model might need more refinement by adding more predictors.
 
@@
 #| warning: false
-#| label: fig-ppcheckandposteriorvsprior
+#| label: fig-pp-check
 #| layout-ncol: 2
 #| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
+
+bayesian_model <-
+  readRDS(file = here::here("models/bayesian_model.rds"))
+
+pp_check(bayesian_model)
 
@@
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
+@tbl-bayesian-model-summary shows us the summary of the bayesian model.
+Since the Rhat value for all parameters are 1, and both Bulk_ESS and Tail_ESS for all parameters are larger than 100, we can conclude that there is excellent convergence in the bayesian model.
 
@@
 #| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
+#| label: tbl-bayesian-model-summary
+#| tbl-cap: "Checking the convergence of the bayesian model using Rhat and ESS"
+
+# Convert the summary to a more structured format
+bayesian_summary <- summary(bayesian_model)
+
+# Extract the relevant part of the summary (coefficients, diagnostics, etc.)
+model_summary_df <- as.data.frame(bayesian_summary$fixed)  # or use `bayesian_summary$summary` for more detailed info
+model_summary_df[] <- lapply(model_summary_df, function(x) {
+  if (is.numeric(x)) round(x, 2) else x
+})
+
+# Create a formatted table with caption
+knitr::kable(model_summary_df, caption = "Checking the convergence of the bayesian model using Rhat and ESS")
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| message: false
+#| warning: false
+#| label: fig-trace-plot
+#| fig-cap: "Checking the convergence of the bayesian model using trace plot"
 #| layout-ncol: 2
+
+plot(bayesian_model)
 

### 2024-12-02 other/datasheet/datasheet_template.qmd
Updated paper and datasheet

---
title: "Datasheet for 'Canadian Grocery Price Data'"
author: Tina Kim
thanks: "Code and data are available at: https://github.com/thk421/Canadian-Grocery-Prices.git."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The dataset was created to drive more competition and reduce collusion in the Canadian grocery sector. Some example tasks in mind were doing an economic analysis of pricing data, visualizing the price of making a standard sandwich at each of the grocers (200g white bread + 20g ham + 20g lettuce...) to find the cheapest vendor, and finding which grocer is generally the cheapest across all comparable product families.

2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - This dataset was collected by Jacob Filipp for Project Hammer.

3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - The project does not have an explicit funder.

4. *Any other comments?*
    - The dataset is still in a developing stage and is actively looking for people to contribute in it.

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- Instances represent individual grocery products sold by Canadian vendors.

2. *How many instances are there in total (of each type, if appropriate)?*
	- There are 14 columns and 12117221 rows across 8 distinct vendors.

3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- This dataset is a sample of products available from major Canadian grocery vendors with a focus on urban areas. Seasonal coverage is limited due to the timeframe of data collection.

4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- Each instance consists of the date and time of data collection, their current and previous prices, price per unit, sale/best-seller/stock status, distinct product id, vendor names, product names, units, and brand names.

5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- No explicit label; the dataset is exploratory, though it supports predictions of affordability and vendor reliability.

6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- Certain entries lack unit sizes or prices due to scraping limitations.

7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- No explicit relationships, though vendor-product relationships and historical trends are implicit.

8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- Splits for exploratory analysis, predictive modeling, and testing may be created based on temporal sequences.

9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- Minor inconsistencies in product matching, redundant data due to a product being listed under multiple categories or being miscategorized, and missing data due to website scraping errors.

10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- The product information was scraped from vendor websites, but the dataset is self-contained.

11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- No confidential data is present.

12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- No offensive data is present.

13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- No direct identification of sub-populations.

14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- No personal data is present.

15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- No, it contains only public product and pricing data.

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- Data was scraped directly from grocery vendors' websites.

2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- Automated web scraping tools were used to collect product listings and prices.

3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- Deterministic sampling, focusing on widely available products across major vendors.

4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- Jacob Filipp from Project Hammer, with no external contributors in the data collecting process.

5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- From February 28, 2024 to the latest load.

6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- No formal reviews, but scraping adhered to legal and ethical guidelines.

7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- Directly through web scraping.

8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- Not applicable, as data is public.

9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- Not applicable.

10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- Not applicable.

11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- No formal impact analysis, but data usage focuses on public benefit.

12. *Any other comments?*
	- Data is limited to the publicly accessible product interface.

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- No.
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- Not applicable.
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- Not applicable.

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- It is not explicitly stated, but the original dataset collecter has encouraged people to play around with the data while suggesting some research questions.

2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- https://jacobfilipp.com/hammer/

3. *What (other) tasks could the dataset be used for?*
	- Predictive modeling, vendor comparison, and price optimization.

4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- Care must be taken to avoid overgeneralizing findings to unrepresented vendors or regions.

5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- Real-time decision-making or extrapolations to non-Canadian markets without validation.

6. *Any other comments?*
	- There is a lot of potential for using this data in sustainability and food security studies.

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- Yes, the goal is for other people to use the dataset to make significant contributions to drive more competition and reduce collusion in the Canadian grocery sector.

2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- The dataset is originally distributed from the website https://jacobfilipp.com/hammer/, but can be distributed through GitHub repositories like https://github.com/thk421/Canadian-Grocery-Prices.git.

3. *When will the dataset be distributed?*
	- Upon publication of the related research.

4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- The dataset will be distributed under the MIT license in the GitHub repository.

5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- No restrictions.

6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- None identified.

7. *Any other comments?*
	- This dataset is accessible to researchers and the public.

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- Jacob Filipp from Project Hammer.

2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- They can be contacted through their website: https://jacobfilipp.com/hammer/

3. *Is there an erratum? If so, please provide a link or other access point.*
	- None currently.

4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- The dataset is constantly being updated with new instances (as of 2024-12-01 20:32:00.43 Eastern Time) and can be checked on the website https://jacobfilipp.com/hammer/.

5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- No, dataset is intended for long-term availability.

6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- Yes, it will be available on the website https://jacobfilipp.com/hammer/, and in case it is not maintained, an older version is archived on the repository https://github.com/thk421/Canadian-Grocery-Prices.git.

7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- Contributions can be made by downloading the data from the website or Contributions via pull requests or issue submissions in the git repository. Validation can be done by contacting the original dataset collector at https://jacobfilipp.com/hammer/.

\newpage

# References

### 2024-12-02 paper/paper.qmd
Updated paper and datasheet

@@
 ---
-title: "Predicting Grocery Price Trends Using Bayesian Models"
-subtitle: "An analysis of price patterns across Canadian grocery vendors to identify the most beneficial vendor for consumers"
+title: "Predicting Canadian Grocery Price Trends Using Bayesian Models"
+subtitle: "An analysis of price patterns across grocery vendors in Canada to identify the most beneficial vendor for consumers"
 author: Tina Kim
@@
 
+@fig-trace-plot shows the traceplots for the bayesian model. [Add more here]
+
 ```{r}

### 2024-12-02 other/datasheet/datasheet_template.qmd
Updated datasheet and references

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "The Canadian Grocery Price Dataset provides a comprehensive collection of product pricing and availability information from major grocery vendors across Canada.
+Designed to support research on affordability, convenience, and sustainability in grocery shopping, the dataset contains detailed entries on data collection dates, vendor names, product names, unit prices, sale statuses, stock statuses, and best-seller statuses.
+Data was collected through automated web scraping, ensuring consistency and temporal relevance, while preprocessing enhanced comparability across vendors.
+This dataset aims to inform consumer decision-making, vendor selection, and policy development by offering insights into pricing trends and vendor performance.
+The dataset is suitable for tasks such as predictive modeling, vendor scoring, and economic analysis, with potential applications in sustainability and food security studies.
+It is shared under an open-source license to encourage broad use and collaboration, with provisions for periodic updates and community contributions."
 format: pdf
@@
 Extract of the questions from @gebru2021datasheets.
+Datset from @grocery_data.
 

### 2024-12-03 scripts/05-exploratory_data_analysis.R
Updated scripts

@@ -11,10 +11,11 @@
 #### Workspace setup ####
 library(tidyverse)
 library(rstanarm)
+library(arrow)
 
 
 #### Read data ####
-analysis_data <- read_csv("data/02-analysis_data/analysis_data.csv")
+analysis_data <- read_parquet("data/02-analysis_data/analysis_data.parquet")
 
 
 #### Analyze data ####

### 2024-12-03 scripts/06-model_data.R

@@ -13,10 +13,11 @@ library(tidyverse)
 library(rstanarm)
 library(brms)
 library(bayesplot)
+library(arrow)
 
 
 #### Read data ####
-analysis_data <- read_csv("data/02-analysis_data/analysis_data.csv")
+analysis_data <- read_parquet("data/02-analysis_data/analysis_data.parquet")
 
 ### Model data ####
 

### 2024-12-03 paper/paper.qmd
Updated paper

@@
 library(rstan)
+library(bayesplot)
 
@@
 # Introduction
-
-Overview paragraph
 
@@
 
-Estimand paragraph
-
 The key estimand of this study is the effect of vendor choice, product category, and temporal trends on grocery pricing across Canada.
-Specifically, we aim to estimate how much more or less consumers might pay for certain product categories when purchasing from different vendors, considering the ongoing trends in grocery pricing and the role that seasonal and economic shifts play.
+Specifically, we aim to estimate how much more or less consumers might pay for certain product categories when purchasing from different vendors, considering the ongoing trends in grocery pricing.
 Through this analysis, we seek to estimate the probability of future price decreases or increases based on these factors.
@@
 
-Results paragraph
-
-The analysis reveals that vendor choice plays a significant role in determining grocery prices, with large national vendors generally offering lower prices compared to local or regional options.
-Furthermore, we observe a clear seasonal trend in grocery pricing, with prices typically rising in the winter months and dropping during certain sales periods.
-The model also indicates that certain product categories, particularly organic and plant-based foods, are subject to higher price volatility, while staple items such as grains and protein foods remain relatively stable.
-
-Why it matters paragraph
+The analysis reveals that vendor choice plays a significant role in determining grocery prices when factoring product categories and temporal attributes.
+NoFrills offers the lowest predicted prices, making it ideal for budget-conscious shoppers, though it has occasional price spikes.
+For those valuing price stability, Loblaws or Metro are better options.
+However, Voila emerges as the top choice, balancing affordability with superior service and reliability.
 
@@
 
-Telegraphing paragraph: The remainder of this paper is structured as follows.
+The remainder of this paper is structured as follows.
 @sec-data describes the data used in the analysis, @sec-model details the Bayesian model and assumptions, @sec-results presents the results, and @sec-discussion offers a discussion of the findings and suggests areas for future research.
+An overview of the sampling methdology is in @sec-appendix-A.
+Additional data details are explained in @sec-appendix-B, and additional model details are noted in @sec-appendix-C.
 
@@
 The tidyverse library [@tidyverse] provides a comprehensive set of tools for data manipulation, cleaning, and visualization, enabling efficient handling of datasets.
-The lubridate library [@lubridate] simplifies working with date-time objects while arrow [@arrow] is used to handle large datasets efficiently, especially for columnar formats like Parquet.
+The lubridate library [@lubridate] simplifies working with date-time objects while arrow [@arrow] is used to handle large datasets efficiently, especially for formats in Parquet.
 For Bayesian modeling, rstanarm [@rstanarm] and brms [@brms] offer powerful tools for fitting and interpreting multilevel models while bayesplot [@bayesplot] aids in the graphical exploration of Bayesian model results, enhancing the interpretation of complex models.
@@
 ## Measurement
-
-Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.
 
@@
 - Another outcome variable is the overall vendor score as seen in @fig-overall-vendor-score, which takes into account multiple criteria to evaluate vendors.
-The highest weight is put on affordability which is measured by the average price per unit of groceries from each vendor.
+The highest weight is put on affordability since it is the primary goal of this paper, and is measured by the average price per unit of groceries from each vendor.
 Some additional factors taken into account are whether products are organic (quality), are in stock (reliability), are on sale or are best sellers (appeal), and if there was a quiet price decrease (transparency).
+The last three factors had the lowest weights because it did not apply to every vendor.
 
@@
 
-- Vendor Name (vendor): There are 8 vendors in this analysis: Voila, Loblaws, NoFrills, Metro, Walmart, TandT, SaveonFoods, and Galleria.
+- Vendor Name (vendor): There are 8 vendors in this analysis as shown in the table below: Voila, Loblaws, NoFrills, Metro, Walmart, TandT, SaveonFoods, and Galleria.
 This variable is crucial since we are comparing results between these vendors.
-For instance, we can factor in the fact that Voila has approximately 4 times the number of products compared to Galleria as seen in the table below.
 
@@
 - Quiet Price Decrease (price_decrease): Tracks subtle pricing adjustments by identifying instances where a product’s price decreased without being labeled as a sale.
-This outcome sheds light on subtle pricing strategies vendors might employ to remain competitive without overt promotional activities.
+This outcome indicates subtle pricing strategies vendors might employ to remain competitive without overt promotional activities.
 It provides insights into whether vendors prioritize transparent discounting (sales) or discreet price reductions.
@@
 
-- Organic Product Status (is_organic): Whether a product is “Organic” serves as a critical predictor for those who want higher quality products while risking affordability.
+- Organic Product Status (is_organic): Whether a product is “Organic” serves as a predictor for those who want higher quality products while risking affordability.
 We compare the availability of high-quality products across vendors and categories by looking for product names that include “fresh” or “organic”.
@@
 
-### Model Justification
+## Model Justification
 
@@
 ## Implementation and Validation  
-The model was implemented in R using the `brms` package. Validation was done by visually checking posterior predictive checks to assess the model's fit.
+The model was implemented in R using the `brms` package.
+Validation was done by visually checking posterior predictive checks to assess the model's fit and is explained with more detail in @sec-appendix-C.
 
@@
 
-The average weekly meal plan cost by vendor and food category in @fig-avg-price-stacked shows in order which vendors overall are the most cost-effective to the least cost-effective.
-By looking at the different colors that represent the food catgory, it can also be compared which food group is more cost-effective in which store.
+The average weekly meal plan cost by vendor and food category in @fig-avg-price-stacked ranks vendors from most to least cost-effective.
+The color-coded food categories also allow for comparisons of which food groups are more cost-effective at specific stores.
 
@@
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+## Paper Summary {#sec-first-point}
+This paper explored several complementary analyses to evaluate grocery pricing trends across vendors and food categories.
+These efforts provided deeper insights into affordability and purchasing strategies aligned with the Canadian Food Guide recommendations.
 
-## Paper Summary {#sec-first-point}
-what was done in this paper?
+Firstly, the average weekly meal plan cost analysis revealed significant differences among vendors, offering a detailed breakdown of pricing trends across food categories.
+This visualization highlighted which vendors were most cost-effective overall and how food group pricing varied between stores.
+The additional breakdown of costs by food group identified which categories—fruits and vegetables, protein, grains, or dairy—had the highest and lowest costs across vendors.
+
+Vendor-specific attributes, such as sale transparency, product appeal, and reliability of being in stock, were also examined as part of a broader evaluation.
+While affordability remained a dominant factor, these additional features provided context for choosing vendors based on consumer priorities beyond cost.
+For example, stores that prioritize organic products or explicitly promote sales emerged as better choices for consumers seeking environmentally conscious or clearly discounted options.
+
+This research also underscored the importance of informed meal planning, where a combination of affordability and quality factors can guide consumer decisions.
+The results have implications for public policy, particularly in addressing food accessibility and encouraging healthier eating habits by promoting cost-effective options for meeting dietary recommendations.
 
@@
 
-According to the predicted future price trends in @fig-future-price-trends, Galleria exhibits relatively stable predicted price trends, with slight variations over time.
-The predicted prices stay consistent, without significant spikes, suggesting steady pricing strategies.
-Loblaws shows stable prices with slight fluctuations but no dramatic increases or decreases.
-This implies consistent pricing practices with minor seasonal or product-related variations.
-Metro displays a similar pattern of stability to Loblaws but with marginally larger variations.
-The periodicity of the vertical bars may indicate uncertainty in the predictions during certain months.
+According to the predicted future price trends in @fig-future-price-trends, most vendors show relatively stable predicted price trends, with slight variations over time.
+However, the periodicity of the vertical bars in the graphs may indicate uncertainty in the predictions during certain months.
+Galleria has a sharp decrease followed by a sharp increase halfway into the year, showing unstability in pricing trends.
+Loblaws and Metro have minor variations and seem overall more stable.
 NoFrills has a noticeable spike mid-period, followed by a return to stability.
@@
 Voila shows a small spike mid-period but generally maintains consistent prices.
-The predicted pricing pattern mirrors a temporary fluctuation, likely tied to vendor or product-specific dynamics.
 Walmart demonstrates the most volatility in its price trends, with noticeable spikes.
@@
 
-@fig-trace-plot shows the traceplots for the bayesian model. [Add more here]
+The traceplots for the bayesian model are shown in three separate figures, @fig-trace-plot-group1, @fig-trace-plot-group2, and @fig-trace-plot-group3 for better visibility.
+These plots display the sampled values for each parameter across iterations for two chains.
+Ideally, the chains should mix well and overlap to indicate convergence.
+Overall, the appearance of the trace plots suggest that the Bayesian model has converged well for all parameters used in the model.
+Both chains overlap and explore the posterior space effectively, with no significant signs of non-convergence.
+This ensures the use of these posterior estimates in the inference and analysis of this paper is appropriate.
 
@@
 #| warning: false
-#| label: fig-trace-plot
-#| fig-cap: "Checking the convergence of the bayesian model using trace plot"
-#| layout-ncol: 2
 
-plot(bayesian_model)
+posterior_samples <- as.matrix(bayesian_model)
 
+# Group 1
+group1_params <- c("b_Intercept", "b_current_price", "b_food_categoryGrainProducts", "b_food_categoryProteinFoods", "b_vendorLoblaws")
+mcmc_trace_group1 <- mcmc_trace(posterior_samples[, group1_params]) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+# Group 2
+group2_params <- c("b_vendorMetro", "b_vendorNoFrills", "b_vendorSaveOnFoods", "b_vendorTandT", "b_vendorVoila")
+mcmc_trace_group2 <- mcmc_trace(posterior_samples[, group2_params]) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+# Group 3
+group3_params <- c("b_vendorWalmart", "b_month", "b_price_decrease", "sd_product_name__Intercept", "sigma")
+mcmc_trace_group3 <- mcmc_trace(posterior_samples[, group3_params]) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| message: false
+#| warning: false
+#| label: fig-trace-plot-group1
+#| fig-cap: "Trace plots for first five predictors"
+
+mcmc_trace_group1
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| message: false
+#| warning: false
+#| label: fig-trace-plot-group2
+#| fig-cap: "Trace plots for sixth to tenth predictors"
+
+mcmc_trace_group2
+```
+
+```{r}
+#| echo: false
+#| eval: true
+#| message: false
+#| warning: false
+#| label: fig-trace-plot-group3
+#| fig-cap: "Trace plots for last five predictors"
+
+mcmc_trace_group3
 ```

### 2024-12-03 scripts/02-download_data.R
Updated scripts

@@ -4,7 +4,8 @@
 # Date: 19 November 2024
 # Contact: tinak.kim@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: None
+# Pre-requisites: 
+  # - The `tidyverse` package must be installed and loaded
 # Any other information needed? None
 
 

### 2024-12-03 scripts/03-clean_data.R

@@ -4,7 +4,9 @@
 # Date: 19 November 2024
 # Contact: tinak.kim@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: Need to have downloaded the data
+# Pre-requisites: 
+  # - Need to have downloaded the data from 02-download_data.R
+  # - The `tidyverse` and `arrow` packages must be installed and loaded
 # Any other information needed? None
 
 #### Workspace setup ####

### 2024-12-03 scripts/04-test_analysis_data.R

@@ -4,15 +4,18 @@
 # Date: 28 November 2024
 # Contact: tinak.kim@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: Have the analysis data loaded
+# Pre-requisites: 
+  # - Have the analysis data loaded
+  # - `tidyverse`, `arrow` and `lubridate` packages must be installed and loaded
 # Any other information needed? None
 
 
 #### Workspace setup ####
 library(tidyverse)
 library(lubridate)
+library(arrow)
 
-analysis_data <- read_csv("data/02-analysis_data/analysis_data.csv")
+analysis_data <- read_parquet("data/02-analysis_data/analysis_data.parquet")
 
 
 #### Test data ####

### 2024-12-03 scripts/05-exploratory_data_analysis.R

@@ -4,7 +4,9 @@
 # Date: 28 November 2024
 # Contact: tinak.kim@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: Have the analysis data loaded
+# Pre-requisites: 
+  # - Have the analysis data loaded
+  # - `tidyverse`, `rstanarm` and `arrow` packages must be installed and loaded
 # Any other information needed? None
 
 

### 2024-12-03 scripts/06-model_data.R

@@ -4,7 +4,10 @@
 # Date: 28 November 2024
 # Contact: tinak.kim@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: Have the 05-exploratory_data_analysis.R script loaded
+# Pre-requisites: 
+  # - Have the 05-exploratory_data_analysis.R script loaded
+  # - `tidyverse`, `rstanarm`, `brms`, `bayesplot` and `arrow` packages 
+      # must be installed and loaded
 # Any other information needed? None
 
 

### 2024-12-03 scripts/00-simulate_data.R
Updated scripts

@@ -4,7 +4,8 @@
 # Date: 19 November 2024
 # Contact: tinak.kim@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: None
+# Pre-requisites:
+  # - The `tidyverse` and `lubridate` packages must be installed and loaded
 # Any other information needed? None
 
 

### 2024-12-03 scripts/02-download_data.R

@@ -27,8 +27,8 @@ unzip(destfile, exdir = "unzipped_files")
 
 # Read the CSV file
 csv_file <- list.files("unzipped_files", pattern = "\\.csv$", full.names = TRUE)
-data_price <- read.csv("unzipped_files/hammer-4-raw.csv")
-data_product <- read.csv("unzipped_files/hammer-4-product.csv")
+data_price <- read_csv("unzipped_files/hammer-4-raw.csv")
+data_product <- read_csv("unzipped_files/hammer-4-product.csv")
 
 
 #### Save data ####
